Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2016-11-13 manuscript.tex
latest

In our view, our density-dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood). In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that resources are spatially localized (e.g. due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition. The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species. We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].

On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$. The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, without arbitrarily assigning an absolute fitness cost to that competition. [citations to sexual selection stuff?]

In the broader context of eco-evolutionary synthesis [? Cite], we can imagine a spectrum of approaches ranging from population genetics, which is entirely evolutionary (no ecology) and has a strong emphasis on genetics, to pure ecological theory with no evolution.

\begin{figure}
\centering
\begin{tabularx}{\linewidth}{lXXXXX}
&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
  Eco-evo:  \\
  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{55}** & \ding{55} \\
  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
\end{tabularx}
\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion''). **As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' or evolutionary suicide.}
\end{figure}

>>>>

In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood). In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that resources are spatially localized (e.g. due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition. The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species. We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].

On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$. The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to linkage disequilibrium between $c$ on the one hand, and $b$ and $d$ on the other. Previous work on this question in the case of sexual conflict used problem-specific fitness models which directly imposed an absolute fitness cost on competition \citep{rankin_2011,vandoorn_2006}.

In Fig. \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here. In a sense this comparison is unfair: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits). As such, Fig. \ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose. Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e. to capture some crucial ecological factors in our most genetically realistic models of evolution. Thus, viewing evolutionary ecology as a spectrum ranging from evolution-only to ecology-only, our model lies closer to the evolution-only end of the spectrum compared to more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology plus mutant invasion --- which lies closer to the ecology-only end of the spectrum.

\begin{figure}
\centering
\begin{tabularx}{\linewidth}{lXXXXX}
&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
  Eco-evo:  \\
  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{55}** & \ding{55} \\
  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
\end{tabularx}
\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology. *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically. Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion''). **As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching''.}
\end{figure}
